### 2016-01-27 hans/bayes-tardis-luminosity-prediction.tex
[hans] Add first Gamma approximations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Bayesian prediction of luminosity distribution based on one simulation run:
% An example from TARDIS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{article}
\usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
\usepackage[numbers]{natbib}

@@

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\refeq}[1]{Eq.~(\ref{#1})}

@@

The Gamma distribution is with shape parameter $\alpha$ and rate
parameter $\beta$ is defined as
\begin{align}
  \lleq{gma}
  p(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
For our purposes, the most important property is that the Gamma distribution is infinitely divisible; i.e., the sum of $N$ Gamma variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
  X_i \sim Gamma(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim Gamma(N \alpha, \beta)
\end{align}

With $\bmphi = (\alpha, \beta)$, the property \refeq{gmb} is
equivalent to
\begin{align}
  \lleq{gmc}
 p(L \cond N, \bmphi) &= \int d\bmL\,\delta(L - \textstyle\sum_j L_j)
  \,p(\bmL\cond N,\bmphi) = Gamma(L \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    p(L\cond n,\bml,m)
  = \frac{1}{p(\bml\cond n)}
  \sum_N p(N\cond n)\, \int d\bmphi\, p(L \cond \bmphi)\, p(\bml\cond m,\bmphi)
  \, p(\bmphi).
\end{align}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Regarding the latter, there is a conjugate prior but
unfortunately the normalization constant can not be evaluated in
closed form. It is thus of no use for us, and we turn to a
noninformative prior. In \cite{Moala:2013}, an overview of possible
choices is given. They conclude that the point estimate given by the
mode of the posterior is nearly insensitive to the prior choice for 30
samples. We require in addition that the posterior is narrow enough so
we group samples in bins to have at least $m=150$ samples in each
bin. The following approximations can be used in different regions of
the parameter space spanned by $n$ and $m$.

\begin{enumerate}
  \item
\end{enumerate}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Bayesian prediction of luminosity distribution based on one simulation run:
% An example from TARDIS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{article}
\usepackage[a4paper, left=20mm, right=20mm, top=25mm, bottom=25mm]{geometry}
\usepackage[numbers]{natbib}
%\usepackage{hyperref}

@@

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\refeq}[1]{Eq.~(\ref{#1})}
\newcommand{\lmax}{\ell_{\rm max}}

@@

The Gamma distribution is with shape parameter $\alpha$ and rate
parameter $\beta$ is defined for $x \geq 0$ as
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} .
\end{align}
$\GammaDist(x \cond \alpha, \beta)$ is defined for $x \in [0, \infty]$
whereas the luminosity of real packets is bounded from above and has a
peak away from the bound. In order to describe the luminosity
distribution of real packets through $\GammaDist(\alpha, \beta)$, we
have to transform the luminosities. To this end, we first take the
largest luminosity from the entire frequency range and increase it
slightly to avoid numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_j \ell_j,
\end{align}
where $ \epsilon=10^{-6}$.
We assume the total number of samples across all bins is large, hence
the effect of considering $\lmax$ a constant independent of the data
is negligible. We then transform the samples as
\begin{align}
  \lleq{gmf}
  x_j \equiv 1- \frac{\ell_j}{\lmax} > 0
\end{align}
and similarly for $L \to X$, then we infer $P(X \cond n, \bmx,
m)$. Only in the end, we transform back to a probability density in
$L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond n, \bml, m) = p(X  \cond n, \bmx, m) \frac{1}{\lmax}
\end{align}
While not obvious at first, a major numerical simplification arises
because of the $L \to X$ transformation. Rewriting the master
equation~(\ref{pmp}) formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond n, \bml, m) = \sum_{N} p(N \cond n) p(L \cond N, \bml, m)
\end{align}
we see that the $ p(L \cond n, \bml, m)$ is a mixture density. If the
distribution of a single luminosity, $p(L \cond N=1, \bml, m)$ is
narrow in the sense that the standard deviation relative to the mean
is at the percent level, then the mixture density has one mode for
each $N$ and all modes are clearly separated. The transformation $L
\to x$ effects a strong overlap between adjacent $p(X \cond N, \bmx,
m)$ and leads to a unimodal $p(X \cond n, \bmx, m)$.
\fred{Show a plot: from hedgehog to single peak}

For our purposes, the most useful property of the Gamma
distribution is infinite divisibility; i.e., the sum of $N$ Gamma
variates is again a Gamma variate
\begin{align}
  \lleq{gmb}
  X_i \sim \GammaDist(\alpha, \beta) \Rightarrow \sum_{i=1}^N X_i \sim \GammaDist(N \alpha, \beta).
\end{align}
With $\bmphi = (\alpha, \beta)$, \refeq{gmb} is equivalent to
\begin{align}
  \lleq{gmc}
 p(X \cond N, \bmphi) &= \int d\bmX\,\delta(X - \textstyle\sum_j X_j)
  \,p(\bmX\cond N,\bmphi) = \GammaDist(X \cond N \alpha, \beta)
\end{align}
and this implies that the master equation~(\ref{pmp}) considerably simplifies as
\begin{align}
  \lleq{gmd}
    p(X\cond n,\bmx,m)
  = \frac{1}{p(\bmx\cond m)}
  \sum_N p(N\cond n)\, \int d\bmphi\, p(X \cond N, \bmphi)\, p(\bmx\cond m,\bmphi)
  \, p(\bmphi).
\end{align}

To explicitly evaluate \refeq{gmd}, we have to overcome the remaining
two difficulties: the sum over $N$ and the integral over
$\bmphi$. Regarding the latter, there is a conjugate prior but
unfortunately the normalization constant can not be evaluated in
closed form. It is thus of no use for us, and we turn to a
noninformative prior. The authors of \cite{moala2013bayesian} present
an overview of possible choices and conclude that the point estimate
given by the mode of the posterior is nearly insensitive to the prior
choice for 30 samples. We require in addition that the posterior is
narrow enough so we group samples in bins to have at least $m=150$
samples in each bin. The following approximations can be used in
different regions of the parameter space spanned by $n$ and $m$.

\begin{enumerate}
  \item For $m$ large, the posterior $p(\bmphi \cond m, \bmx) =
  p(\bmx\cond m,\bmphi) \, p(\bmphi) / p(\bmx \cond m)$ approaches a
  multivariate Gaussian with variance tending to zero as $1/m$ in the
  limit $m \to \infty$.  \fred{check} We can then assume the
  parameters $\bmphi$ known and fix them at the posterior mode
  $\bmphi^*$; i.e., $p(\bmphi \cond m, \bmx) \to \delta(\bmphi -
  \bmphi^*)$.
  \item For $n$ large, the Negative Binomial $p(N \cond n)$ approaches
  a Gaussian, defining a region around $N=n$ with the dominant
  contribution to the sum over $N$ in \refeq{gmd}. The central limit
  theorem states that for large $N$, $P(X \cond N, \bmphi) \approx
  \mathcal{N}(X \cond N \, E_\bmphi[X_j], N \, V_\bmphi[X_j])$. For
  $m$ large enough, we could then plug in the sample-mean and
  sample-variance estimates and bypass the assumption of a Gamma
  distribution altogether.
\end{enumerate}
